Relationships in the extremes and their influence on competition and coexistence: Supporting information

### 2022-03-10 paper/SupMat.Rnw
finished editing S10 and deleted S11 of the sup mat

%Defintion of E and C
Our decomposition of GWR is defined by the choices $E_i(t)=\theta(t)$, 
$C_{i}(t)=(K_i(t)+S(t))/S(t)$, which then imply $r_i(E_i,C_{i})=\frac{V_i(E_i)}{C_{i}} -D$. 
This choice follows one of the choices made in the R computer code provided by \citep{Ellner_2016,Ellner_2019}.
Other choices are possible. For instance, in addition to this choice, 
\cite{Ellner_2016} also considers an alternative choice $E_i(t)=V_i(t)$,
as well as some other possibilities discussed in their supporting materials and 
represented in their computer code.

%outline how the quantities E_i and C_i\i were obtained from simulations. 
To obtain $E_i$ and $C_{i \bs i}$, the competition model was simulated for $T=3000$ time steps.
Environmental temperature fluctuations were defined by the paramters $a$, $P$, and $\theta_0$ using 
$\theta(t)=asin(2\pi t/P)+\theta_0$, and this determined $E_i(t)$ for $i=1,2$. To obtain 
$C_{i \bs i}$, a simulation was carried out starting from $x_i=0$ and $x_j=20$.
Dilution was set to $D=0.09$ and inflow silicate concentration was set at $S_0=35$. A burn in time 
of $\frac{2}{3}T$ was cut from the simulation and the $S(t)$ time series from the remainder of the simulation,
along with $K_i(t)=K_i(\theta(t))$ were plugged in to the expression $C_{i}(t)=(K_i(t)+S(t))/S(t)$.

@@

The decomposition of storage effects and quantification of contributions of ATAs depends on 
removing asymmetric tail associations from the bivariate random variable
$(E_i,C_{i \bs i})$, while maintaining its correlation and marginal distributions, to produce a new
random variable $(E_i^\nat,C_{i \bs i}^\nat)$.
In the Theory section of the main text, we decribed this removal process
when the distributions of $E_i$ and $C_{i \bs i}$ are known exactly, but for the 
diatom example explored for goal G4 of the main text (see also SI section \ref{sect:diatom_model}), 
we only had samples from $(E_i,C_{i \bs i})$, based on simulations. We here describe how to
produce a sample from $(E_i^\nat,C_{i \bs i}^\nat)$ in this modified context. 
Let $(E_{i,k},C_{i \bs i,k})$, $k=1,\ldots,T$ denote the samples from $(E_i,C_{i \bs i})$.
Let $\rk(E_{i,k})$ be the rank of $E_{i,k}$ in the set 
of $E_{i,k}$, $k=1,\ldots,T$, where the smallest value has rank $1$. Likewise 
define $\rk(C_{i \bs i,k})$ as the rank of $C_{i \bs i,k}$ in the set
of $C_{i \bs i,k}$, $k=1,\ldots,T$. Define the ``normalized rank''
$\nrk(E_{i,k})$ as $\rk(E_{i,k})/(T+1)$, and likewise define 
$\nrk(C_{i \bs i,k})$ as $\rk(C_{i \bs i,k})/(T+1)$. Letting $\varphi$ be the CDF
of a standard normal distribution, the samples $\varphi^{-1}(\nrk(E_{i,k}))$, $k=1,\ldots,T$
are standard normally distributed, as are the samples 
$\varphi^{-1}(\nrk(C_{i \bs i,k}))$, $k=1,\ldots,T$. Let $\rho_i$ be the covariance 
of these samples. We then define a bivariate normal random variable 
$(e_i,c_{i \bs i})$ such that $\E(e_i)=\E(c_{i \bs i})=0$, $\var(e_i)=\var(c_{i \bs i})=1$,
and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We used $\ddot{\rho}_i = \rho_i$.
Generating a sample $(e_{i,k},c_{i \bs i,k})$, $k=1,\ldots,T$ from this distribution,
we permute the ordered set $(E_{i,1},E_{i,2},\ldots,E_{i,T})$ so that the ranks of the 
permuted ordered set equal the ranks of the corresponding elements of the ordered
set $(e_{i,1},e_{i,2},\ldots,e_{i,T})$. We likewise permute the ordered set 
$(C_{i \bs i,1},C_{i \bs i,2},\ldots,C_{i \bs i,T})$ so the ranks of the permuted result
match the ranks of $(c_{i \bs i,1},c_{i \bs i,2},\ldots,c_{i \bs i,T})$.
This step works as long as there are no ties among the $E_{i,k}$ or among
the $C_{i \bs i,k}$. There were no ties when we applied these methods to the diatom system. 
The two permuted ordered sets, here denoted $(E_{i,1}^\nat,E_{i,2}^\nat,\ldots,E_{i,T}^\nat)$
and $(C_{i \bs i,1}^\nat,C_{i \bs i,2}^\nat,\ldots,C_{i \bs i,T}^\nat)$,
when combined to make ordered pairs, form the desired sample from 
$(E_i^\nat,C_{i \bs i}^\nat)$. Because $(E_{i,1}^\nat,\ldots,E_{i,T}^\nat)$
is a permutation of $(E_{i,1},\ldots,E_{i,T})$ and 
$(C_{i \bs i,1}^\nat,\ldots,C_{i \bs i,T}^\nat)$ is a permutation of
$(C_{i \bs i,1},\ldots,C_{i \bs i,T})$, marginal distributions of our
sample from $(E_i^\nat,C_{i \bs i}^\nat)$ were the same as those of our original 
sample from $(E_i, C_{i \bs i})$. The choice of bivariate normal
$(e_i,c_{i \bs i})$ ensures that our resulting sample shows symmetric tail association.
Finally, we caution that sampling variation means any quantity computed from our resulting sample 
$(E_{i,k}^\nat,C_{i \bs i,k}^\nat)$, $k=1,\ldots,T$ should be recomputed for 
multiple such samples, all computed via the same process, and a median of 
the resulting distribution of values should be used.

\section{Fig. \ref{MT-fig:fig4} does not depend on the value of $q_{ij}$ used}\label{sect:F4andqij}

%***DAN: Jasmin mentioned she had some math proving this, it needs to be added here.
%It has been added but Dan needs to edit it.

In Fig. \ref{MT-fig:fig4} we plot IGR , $\Delta_i^{[EC]}$, and IGR $-\Delta_i^{[EC]}$ (see equation \ref{eq:decomp_overall_delta}) and the text below it.  
From section \ref{sect:math_simplifications}, 
\begin{align}
\olr_{j \bs i} &= \E[\ln(1-\delta+E_j/C_{j \bs i})] \\
&= \E[\ln(1-\delta+\delta B_j/B_j)] \\
&= 0,
\end{align}
and
\begin{align}
\olep_j^{[EC]} &= \E[r_j(E_j,C_{j \bs i})] - \E[r_j(E_j^\nat , C_{j \bs i}^\nat)] \\
&= \olr_{j \bs i} - \olr_{j \bs i}^\nat \\
&= 0. %\label{eq:decomp:j:EnatC}
\end{align}
Then $q_{ij}$ does not matter for IGR or $\Delta_i^{[EC]}$ because the second term in $\olr_{i \bs i}-q_{ij}\olr_{j \bs i}$ and $\Delta_i^{[EC]}=\olep_i^{[EC]}-q_{ij} \olep_j^{[EC]}$ are zero so 
\begin{equation}
\olr_{i \bs i}-q_{ij}\olr_{j \bs i}=\olr_{i \bs i}
\end{equation}
and 
\begin{equation}
\Delta_i^{[EC]}=\olep_i^{[EC]}-q_{ij} \olep_j^{[EC]}=\olep_i^{[EC]}
\end{equation}
Similarly for IGR $-\Delta_i^{[EC]}$, 
\begin{equation}
(\olr_{i \bs i}-q_{ij}\olr_{j \bs i})-(\olep_i^{[EC]}-q_{ij} \olep_j^{[EC]}) = \olr_{i \bs i}-\olep_i^{[EC]}.
\end{equation}
Thus $q_{ij}$ does not matter for anything computed for Fig. \ref{MT-fig:fig4}.

\section{     \emph{Cyclotella psedosteligera} as the invader}\label{sect:Invader2}
We generated plots analougous to Fig. \ref{MT-fig:fig5} and Fig. \ref{MT-fig:fig6} for the reverse invader-resident scenario where \emph{C. pseudosteligera} is the invader and \emph{F. crotonensis} is the resisent to compute the IGR and decompose coexistence mechanisms of \emph{C. pseudosteligera}. Fig. \ref{fig:fig5_2} is analogous to Fig. \ref{MT-fig:fig5} and shows changes in the IGR and decomposed mechanisms in response to change in one temperature parameter while the other two are held constant. Within the parameters of our simulations, the IGR of \emph{C. pseudosteligera}

>>>>

%Defintions of E and C
Our decomposition of GWR is defined by the choices $E_i(t)=\theta(t)$, 
$C_{i}(t)=(K_i(t)+S(t))/S(t)$, which then imply $r_i(E_i,C_{i})=\frac{V_i(E_i)}{C_{i}} -D$. 
This choice follows one of the choices made in the R computer code provided by \citep{Ellner_2016,Ellner_2019}.
Other choices are possible. For instance, in addition to this choice, 
\cite{Ellner_2016} also considers an alternative choice $E_i(t)=V_i(t)$,
as well as some other possibilities discussed in their supporting materials and 
represented in their computer code.

%Outline how the quantities E_i and C_i\i were obtained from simulations. 
To obtain $E_i$ and $C_{i \bs i}$, the competition model was simulated for $T=3000$ time steps.
Environmental temperature fluctuations were defined by the paramters $a$, $P$, and $\theta_0$ using 
$\theta(t)=asin(2\pi t/P)+\theta_0$, and this determined $E_i(t)$ for $i=1,2$. To obtain 
$C_{i \bs i}$, a simulation was carried out starting from $x_i=0$ and $x_j=20$.
Dilution was set to $D=0.09$ and inflow silicate concentration was set at $S_0=35$. A burn in time 
of $\frac{2}{3}T$ was cut from the simulation and the $S(t)$ time series from the remainder of the simulation,
along with $K_i(t)=K_i(\theta(t))$ were plugged in to the expression $C_{i}(t)=(K_i(t)+S(t))/S(t)$.

@@

%***DAN: Dan still needs to edit this section
The decomposition of storage effects and quantification of contributions of ATAs depends on 
removing asymmetric tail associations from the bivariate random variable
$(E_i,C_{i \bs i})$, while maintaining its correlation and marginal distributions, to produce a new
random variable $(E_i^\nat,C_{i \bs i}^\nat)$.
In the Theory section of the main text, we decribed this removal process
when the distributions of $E_i$ and $C_{i \bs i}$ are known exactly, but for the 
diatom example explored for goal G4 of the main text (see also SI section \ref{sect:diatom_model}), 
we only had samples from $(E_i,C_{i \bs i})$, based on simulations. We here describe how to
produce a sample from $(E_i^\nat,C_{i \bs i}^\nat)$ in this modified context. 
Let $(E_{i,k},C_{i \bs i,k})$, $k=1,\ldots,T$ denote the samples from $(E_i,C_{i \bs i})$.
Let $\rk(E_{i,k})$ be the rank of $E_{i,k}$ in the set 
of $E_{i,k}$, $k=1,\ldots,T$, where the smallest value has rank $1$. Likewise 
define $\rk(C_{i \bs i,k})$ as the rank of $C_{i \bs i,k}$ in the set
of $C_{i \bs i,k}$, $k=1,\ldots,T$. Define the ``normalized rank''
$\nrk(E_{i,k})$ as $\rk(E_{i,k})/(T+1)$, and likewise define 
$\nrk(C_{i \bs i,k})$ as $\rk(C_{i \bs i,k})/(T+1)$. Letting $\varphi$ be the CDF
of a standard normal distribution, the samples $\varphi^{-1}(\nrk(E_{i,k}))$, $k=1,\ldots,T$
are standard normally distributed, as are the samples 
$\varphi^{-1}(\nrk(C_{i \bs i,k}))$, $k=1,\ldots,T$. Let $\rho_i$ be the covariance 
of these samples. We then define a bivariate normal random variable 
$(e_i,c_{i \bs i})$ such that $\E(e_i)=\E(c_{i \bs i})=0$, $\var(e_i)=\var(c_{i \bs i})=1$,
and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We used $\ddot{\rho}_i = \rho_i$.
Generating a sample $(e_{i,k},c_{i \bs i,k})$, $k=1,\ldots,T$ from this distribution,
we permute the ordered set $(E_{i,1},E_{i,2},\ldots,E_{i,T})$ so that the ranks of the 
permuted ordered set equal the ranks of the corresponding elements of the ordered
set $(e_{i,1},e_{i,2},\ldots,e_{i,T})$. We likewise permute the ordered set 
$(C_{i \bs i,1},C_{i \bs i,2},\ldots,C_{i \bs i,T})$ so the ranks of the permuted result
match the ranks of $(c_{i \bs i,1},c_{i \bs i,2},\ldots,c_{i \bs i,T})$.
This step works as long as there are no ties among the $E_{i,k}$ or among
the $C_{i \bs i,k}$. There were no ties when we applied these methods to the diatom system. 
The two permuted ordered sets, here denoted $(E_{i,1}^\nat,E_{i,2}^\nat,\ldots,E_{i,T}^\nat)$
and $(C_{i \bs i,1}^\nat,C_{i \bs i,2}^\nat,\ldots,C_{i \bs i,T}^\nat)$,
when combined to make ordered pairs, form the desired sample from 
$(E_i^\nat,C_{i \bs i}^\nat)$. Because $(E_{i,1}^\nat,\ldots,E_{i,T}^\nat)$
is a permutation of $(E_{i,1},\ldots,E_{i,T})$ and 
$(C_{i \bs i,1}^\nat,\ldots,C_{i \bs i,T}^\nat)$ is a permutation of
$(C_{i \bs i,1},\ldots,C_{i \bs i,T})$, marginal distributions of our
sample from $(E_i^\nat,C_{i \bs i}^\nat)$ were the same as those of our original 
sample from $(E_i, C_{i \bs i})$. The choice of bivariate normal
$(e_i,c_{i \bs i})$ ensures that our resulting sample shows symmetric tail association.
Finally, we caution that sampling variation means any quantity computed from our resulting sample 
$(E_{i,k}^\nat,C_{i \bs i,k}^\nat)$, $k=1,\ldots,T$ should be recomputed for 
multiple such samples, all computed via the same process, and a median of 
the resulting distribution of values should be used.

\section{      Fig. \ref{MT-fig:fig4} does not depend on the value of $q_{ij}$ used}\label{sect:F4andqij}

In Fig. \ref{MT-fig:fig4} of the main text we plotted GWR , $\Delta_i^{[EC]}$, and $\text{GWR}-\Delta_i^{[EC]}$ 
(see equation \ref{eq:decomp_overall_delta} and the text below it) for the lottery model. We here show
the resulting plot does not depend on which of the two choices for the scaling factors $q_{ij}$ is used.
But $\text{GWR} = \olr_{i \bs i} - q_{ij} \olr_{j \bs i}$ and $\Delta_i^{[EC]}=\olep_i^{[EC]}-q_{ij}\olep_j^{[EC]}$,
so it suffices to show $\olr_{j \bs i}=0$ and $\olep_j^{[EC]}=0$. The first of these is known true, \emph{a priori}, 
and the second was demonstrated in section \ref{sect:math_simplifications}.
